### 2017-10-02 chapters/paper4/main.tex
Incorporate Sams suggestions

  Finite element models of human left ventricles (LV) of one healthy
  case and one patient diagnosed with left bundle branch block (LBBB)
  were fitted to clinical data consisting of LV volume, regional
  longitudinal strain, and non-invasive estimates of LV pressure. Excellent
  fit between measured and simulated observations was obtained by means
  of gradient-based minimization of model-data mismatch.

  Regional myocardial work was assessed using several approaches,
  including the area of pressure-strain loops and the area of fiber stress-strain
  loops. Qualitative indices of efficiency and quantitative measures of
  mechanical power density were compared using the different approaches,
  including the well established pressure-volume area.

  Results show that the different approaches for computing regional
  work are qualitatively similar, and are able to capture regional
  dysfunction. However, the different approaches are quantitatively
  very different, and total mechanical work computed using the finite
  element model are in close agreement with the actual stroke work.

  In conclusion, mechanical work computed using a personalized cardiac
  mechanics model is quantitatively similar to the actual mechanical
  work estimated by the area of the pressure-volume loop. This allows
  for potentially non-invasive quantitative assessment of regional
  myocardial work, and would greatly improve diagnostics of patients
  with regional dysfunction, such as dyssynchronous heart failure.

@@

Typically for LBBB is that septal segments are activated early,
causing a stretch of the opposing lateral wall. The 
contraction of the late-activated lateral wall then results in a 
lengthening of the septal segments during systole, which can be
identified as wasted work. Recent studies 
\cite{vecera2016wasted} have shown that wasted septal work due to the
uncoordinated contraction might be a better predictor for response to
CRT, and that the amount of wasted work reflects the potential for
recovery. In these calculations, regional work is computed as the area
of the pressure-strain loops, where regional longitudinal strain is
measured using speckle tracking echocardiography (SPE), and left
ventricular pressure (LVP) is estimated using a non-invasive method
\cite{russell2012novel}. The same approach, though with other ways of
assessing regional strain and LVP, has been used in earlier studies to
quantify regional myocardial work
\cite{tyberg1974analysis,theroux1974regional}. These calculations are
inspired by the well established correlation between pressure-volume
(PV) area and myocardial oxygen consumption \cite{suga1979total}.
%  the
% area bounded by the end-systolic PV line, the systolic limb of the PV
% loop trajectory and the volume-axis,

@@

Although the pressure-strain loop area as an estimate of regional
myocardial work has shown to be a useful metric in e.g predicting the
response to CRT, it suffers from limitations when compared to computation of 
``real'' myocardial work. Intraventricular pressure is
used as a surrogate for stress, because stress is not possible to
directly measure in an intact heart
\cite{huisman1980measurement}. Consequently, in these calculations all
segments are assumed to be subjected to the same stress, regardless of
differences in overall size, curvature or wall thickness.

% These limitations have been evaluated by Skulstad et al
% \cite{skulstad2002postsystolic}, who computed regional work
% based on longitudinal strain and estimated wall stress using the law of
% Laplace, which accounts for the wall thickness and radius of
% curvature. They found essentially the same results when compared to
% estimates using the cavity pressure as surrogate for stress.

These limitations have been evaluated by Delhaas et. al
\cite{delhaas1994regional} who showed that
the area of regional fiber stress-strain loops was correlated to
regional oxygen demand in six canine left ventricles. In this study,
regional fiber stress was estimated based on assumptions of rotational
symmetry \cite{arts1991relation}, with each segment scaled according to
measured fiber strain. Fiber strain was measured at the subepicardial
layers using a two-dimensional video technique.

However, several simplifications, in particular regarding symmetry, is
made in these simplified models \cite{zhang2011comparison}, and the
measurement has to be done {\it in situ}.

@@

However, a framework for assessing regional estimates of myocardial
work, for which, when all regions are considered, sums up to the total
stroke work needed to eject blood to the body, is still
lacking. Furthermore, in order to be clinically useful, the framework
has to be accurate in terms of being able to replicate the mechanical
conditions observed clinically, and fast in terms of computing these
estimates within a reasonable time frame.

In this study we apply a recently developed framework
\cite{balaban2017high} for constraining a mechanical model to clinical
data. This framework employ gradient-based optimization to estimate high
dimensional parameters in order to minimize the model-data mismatch,
and therefore offers the needed accuracy. Moreover, the framework
utilizes state-of-the-art numerical methods which computes the
gradient efficiently by solving the corresponding adjoint equation,
and thereby provides these estimates within a reasonable time frame.

@@

Incompressibility is forced by using a two-field variational approach
where the term
\begin{align}
\Psi_{\text{vol}} = - p (J-1), 
\end{align}
is added to the total strain energy density, and the variable $p$ is
introduced as a Lagrange multiplier.

@@

The force balance equation can be derived from the classical Cauchy
momentum equation, and with the assumption that the inertial term and
external body forces are negligible, the strong formulation yields
\begin{align}
  \nabla \cdot \FPiola &= 0, \\
  J &= 1, 
\end{align}
with $\FPiola = \frac{\partial \Psi}{\partial \F}$ being the first
Piola-Kirchhoff stress tenors.

To solve the problem we employ the Galerkin finite element method, which is based
on choosing proper finite dimensional subspaces $V_h \subset [H_0^1(\Omega)]^3$ and
$Q_h \subset L^2(\Omega)$ and solve the weak formulation of the problem. In
the Lagrangian form, the weak form reads: {\it find $(\uvec,p)
\in V_h\times Q_h$ such that for all variations $(\delta
\uvec, \delta p) \in V_h \times Q_h $ and $\left. \left( \uvec \cdot \N \right)
\right|_{\partial \Omega_{\mathrm{base}}} = 0$ }:
\begin{align}
\begin{split}
  F(\uvec, p;\delta \uvec, \delta p) =
  &\int_{\Omega} \FPiola :  \Grad \delta \uvec  \mathrm{d} V  \\
  &+\int_{\Omega} p J \F^{-T} : \Grad \ \delta \uvec   \mathrm{d} V  
  +\int_{\Omega} (J - 1) \delta p \mathrm{d} V \\
  &+\int_{\partial \Omega_{\text{base}}} k \uvec \cdot \N \mathrm{d}S
  - \int_{\partial \Omega_{\text{endo}}} \plv J \F^{-T} \N \cdot \delta \uvec \mathrm{d}S= 0.
  \label{eq:force_balance}
\end{split}
\end{align}
Here, the ventricular domain is anchored by constrained the basal
motion by a linear spring-term with stiffness $k$, and the
endocardial cavity pressure $\plv$ is enforced as a Neumann boundary
condition on the endocardium.

@@

Work can be computed by considering any of the
work conjugate stress-strain pairs \cite{holzapfel2000nonlinear}. In
the current configuration it is common to use the Cauchy stress tensor
$\Cauchy = \FPiola \F^T$ and the Almansi strain tensor $\mathbf{b} =
\frac{1}{2} \left(\I -  \F^{-T} \F^{-1} \right)$, while in
the reference configuration we use the second Piola-Kirchhoff stress
tensor $\SPiola = \F^{-1} \FPiola$, and the Green-Lagrange strain
tensor $\E = \frac{1}{2}\left(\C - \I \right)$. Thus, in the Lagrangian
description, the total mechanical work from time $t_1$ to $t_2$ in a
region $\Omega_j$ is given by
\begin{align}
   \mathcal{W}(t_1, t_2)[\Omega_j] = 
  -\int_{t_1}^{t_2} \int_{\Omega_j}\SPiola(t, \Xvec)
  : \dot{\E}(t, \Xvec) \mathrm{d}\Xvec \mathrm{d}t, 
\end{align}
where $ \dot{\E} $ is the rate of the Green-Lagrange strain tensor.
By dividing by the volume of the region of interest $| \Omega_j |$,
we obtain the average work density in that region. Note the sign
convention used here illustrates that a negative strain-rate
represents positive work. To only compute the
work along a specific direction, e.g along the fibers, the full stress
and strain tensors are simply interchanged with their respective
components in that direction, and the double contraction of the tenors
is changed to a scalar multiplication.

The mechanical work needed to eject blood from time $t_1$ to
$t_2$ is given by
\begin{align}
   \mathcal{W}(t_1, t_2)[\Omega] = 
  -\int_{t_1}^{t_2} \frac{\mathrm{d}V}{\mathrm{d}t} P(t) \mathrm{d}t,
\end{align}
where $V(t)$ and $P(t)$ refers respectively to the cavity volume and pressure in
the left ventricle $\Omega$ at time $t$. If the $t_1$ and $t_2$
represents respectively the start and end of the cycle, then normalizing with respect to the
total ventricular wall volume  $| \Omega |$, would yield a measure of
total mechanical work density (TMWD). Diving the TMWD by the total
cycle time, would yield a measure of total mechanical power
density (TMPD). We note that for analysis of performance, TMPD is
a better measure than total work because it is independent of
morphology and heart rate.

\subsubsection{Quasi-static approximation}
The mechanical model described in Section
\ref{sec:mechanical_modeling} is quasi-static, meaning that the
inertial forces are assumed to be so small that the
equilibrium equation \eqref{eq:force_balance} holds at any discrete
time point. Therefore, strain rate is estimated using a
backward difference approximation \cite{wang2011myocardial}.
Suppose we have $N$ discrete time points $t = 0,1, \cdots N$, where $ t = 0$
refers to the the reference configuration, and let $\SPiola(t, \Xvec)$
and $\E(t, \Xvec)$ be respectively the stress and strain tensors at
time $t$ and position $\Xvec$ in the reference configuration. In the
following, the full stress and strain tensors are used, but the methodology
also applies to any component of the stress and strain tensor. We define
\begin{align}
  \overline{\SPiola}(t,\Xvec) &= \frac{1}{2}\left( \mathbf{S}(t,\Xvec) + \mathbf{S}(t-1,\Xvec) \right),\\
  d\E(t,\Xvec) &= \E(t,\Xvec) - \E(t-1,\Xvec),
\end{align}
for $t = 1,\cdots, N$, and note that $ \mathbf{S}(0) = \mathbf{E}(0) = \mathbf{0}$.
The mechanical work in a region $\Omega_j$ from time $t=m$ to time $t=n$ can now be approximated
as 
\begin{align}
  \begin{split}
   \mathcal{W}(m, n)[\Omega_j]
  &= -\int_{t=m}^{t=n} \int_{\Omega_j}\SPiola(t, \Xvec) : \dot{\E}(t, \Xvec) \mathrm{d}\Xvec \mathrm{d}t \\
  &\approx - \sum_{t = m}^{n}\int_{\Omega_j}\overline{\SPiola}(t, \Xvec) : d\E(t, \Xvec) \mathrm{d}\Xvec
  \end{split}
\label{eq:total_work}
\end{align}
Similarly, the mechanical work computed from the pressure-volume loop area can be estimated as  
\begin{align}
  \begin{split}
  \mathcal{W}(m,n)[\Omega]
  &= - \int_{t=m}^{t=n} \frac{\mathrm{d}V}{\mathrm{d}t} P(t) \mathrm{d}t \\
  &\approx -\sum_{t = m}^{n}  (V(t) - V(t-1)) \cdot \frac{P(t) + P(t-1)}{2}.
  \end{split}
\end{align}

% Note that calculation of segmental work in Reference
% \cite{russell2013assessment} is exactly the same but with pressure
% used as a surrogate for stress and longitudinal strain used in stead of
% fiber strain.

@@

During systole, most segments shorten and performs positive work
($\mathcal{W}_{\mathrm{pos}}$). However some segments might elongate,
resulting in energy loss, and we say that these segments do negative work
($\mathcal{W}_{\mathrm{neg}}$). Positive work is identified as the points with
$d\E_{\ef}(t) < 0$ (shortening), while negative work is identified as points with
$d\E_{\ef}(t) > 0$ (lengthening).
A local measure of the energetic efficacy of each segment is the ratio between the
negative and positive work, an index known as the segmental wasted
work ratio (WWR) \cite{russell2013assessment},
\begin{align}
  \mathrm{WWR} = \frac{|\mathcal{W}_{\mathrm{neg}}|}{\mathcal{W}_{\mathrm{pos}}}
\end{align}
Another measure of efficiency, formulated by Niederer et al
\cite{niederer2009role} is the ratio of positive work and the total
work
\begin{align}
  \eta = \frac{\mathcal{W}_{\mathrm{pos}}}{\mathcal{W}_{\mathrm{pos}} + |\mathcal{W}_{\mathrm{neg}}|}.
  \label{eq:efficiency}
\end{align}

The two indices are related as follows: $\eta = \left( 1 +
  \mathrm{WWR} \right)^{-1}$, and we will therefore only report values
of the efficiency \eqref{eq:efficiency}, for which a value close to 1.0 indicate that most
of the performed work is positive. Since we expect work during systole
to be the main contribution to the overall ejection of blood, this
index is computed by only including work done from mitral valve closing to
mitral valve opening \cite{russell2013assessment}.

Note that with the current sign-convention, negative work is by
definition negative, which is why we take the absolute value of the
negative work.

@@

In this study 4D echocardiographic images of one patient diagnosed
with left bundle branch block (LBBB) and one healthy volunteer was
acquired using a GE Vingmed E9 machine (GE Healthcare Vingmed, Horten,
Norway). 
The analysis of the echocardiographic images was conducted with the
GE software package EchoPac version 202. Longitudinal strain was
measured via speckle tracking echocardiography (SPE) in the 17 AHA segments
\cite{cerqueira2002standardized}. However, the apical segments for both
subjects as well as the inferior and posterior basal segment for the
normal case were excluded due to poor quality.

@@

At each point the optimal value of the previous point is used as
initial guess, and zero for the first point of active contraction.

\section{Results}
For the normal and LBBB case defined above, we ran the
full pipeline on two meshes, where one was a refined version of the
other. For the normal case, the original and refined mesh had
respectively 2569 and 5258 elements, while the LBBB case
had 2572 elements in the original mesh and 5114 elements in the
refined mesh. Results on the original and refined geometries gave
close to identical results, and we will therefore only present the
results using the coarsest geometry.

@@

During the estimation of the active and passive parameters outlined in Section
\ref{sec:active_param_estim} and \ref{sec:passive_param_estim}
respectively, the optimization was terminated if the value between two
successive iterations was less then $10^{-6}$ or the number of
iterations exceeded 100. For the material parameter, an upper and lower
bound was set to $50$ and $0.05$ kPa respectively, while for the active
parameter these bounds were set to $0.4$ and $0.0$, allowing for up to
40 $\%$ active shortening of the myofibers.

@@

The total mechanical power density (TMPD) is also computed using four
different approaches: 1) area of the pressure-volume loop normalized
to ventricular wall volume, 2) using the full stress and strain tensors,
3) using the fiber component of the stress and strain tensors, and 4)
using global longitudinal strain and estimated pressure. Table
\ref{tab:total_mechanical_power} show these values for the two
patients, in units kW m$^{-3}$. In the calculations of pressure volume
area, the simulated volume is used, with the unloaded volume and zero
pressure included as reference point.

@@

A qualitative comparison of the pressure-strain loops and fiber
stress-strain loops are illustrated in Figure
\ref{fig:regional_work_indices}, where we plot the regional efficiency
\eqref{eq:efficiency} using the two measures of work. We note that the
efficiency computed using the fiber stress-strain loops are
lower than for the pressure-strain loops but show the same trend, i.e
that regions with reduced efficiency coincide.

@@

The values in Table \ref{tab:total_mechanical_power} indicate that the
total mechanical work performed in the direction of the fibers is
approximately 60 $\%$ of the 
total mechanical work computed by taking into account the full stress
and strain tensor (LBBB: 60.5 $\%$, normal:58.7 $\%$). Of course, the
amount of work performed in transverse direction of the fibers will depend upon the
particular form of the active deformation gradient in
\eqref{eq:active_strain_Fa_gjerald}. In the active stress approach on 
the other hand, where the total Cauchy stress tensor is additively decomposed into
active and passive contributions, it is common to add an active stress
in the transverse direction, which is also reported
experimentally \cite{lin1998multiaxial}. By assuming
a volume preserving active deformation gradient, the amount of
work in the cross fiber direction as computed here is about $40 \%$, a
value that has been used in other modeling studies using the active
stress approach \cite{sun2009computationally}.

The values of regional TMPD in the mid segments presented
in Figure \ref{fig:regional_power_density} show that the septal
segments for the LBBB patient actually consumes energy and yield a
negative TMPD. On the other hand, the healthy heart shows positive TMPD
in all segments and generally higher values than the LBBB heart.
Since the TMPD computed using stress and strain derived from the
underlying constitutive equation agrees very well with the TMPD
computed from the PV area there is reason to believe that the regional
TMPD reflects the actual regional performance, and future studies should
investigate how well these values correlate with regional myocardial
oxygen consumption.

The TMPD computed using the pressure and global longitudinal strain loop
only accounts for a small fraction of the total mechanical power.
This indicates that quantitatively this is not a good measure of
myocardial work. However, the area of the pressure-strain loops have shown to be a good metric
of regional efficiency and a promising index for predicting response
to CRT \cite{vecera2016wasted}. From Figure
\ref{fig:regional_work_indices} we see the same trends in terms of efficiency and wasted
work can be seen regardless of how work is computed, namely that early
activated segments are less efficient and have a higher wasted work
ratio than the other segments for the LBBB case. Therefore, relative
comparison of segments using pressure-strain loops seems to give the
same picture as when using the fiber stress-strain loops. 
The pressure strain loops for the LBBB case all have an
anti-clockwise orientation, indicating generation of work in all
segments. On the other hand, the fiber stress-strain loops in the
septal segments show a clockwise orientation which is evidence of
work usage. In other words, the two approaches show the same
trend, but the effect of the wasted work is magnified in the fiber
stress-strain loops.

@@

The optimization problem during the active phase is highly
under-constrained, which is why we apply a Tikhonov regularization to
restrict the parameter space, so that smooth value of the controls are
favored in the optimization. While the main motivation for this choice
is numerical stability, whether this is valid assumption from a
physiological point of view could be debated, especially in the case
of ventricular dyssynchrony.

Experiments have shown both a visco-elastic and an orthotropic
behavior of the myocardium
\cite{dokos2002shear,sommer2015biomechanical}.
Visco-elasticity would certainly play a role in myocardial work estimation,
as the energy is preserved in a hyperelastic material while some energy dissipate when
taken into account visco-elasticity. However, the effect of
visco-elasticity and orthotropy is not modeled here, and remains the
subject for future studies.

>>>>

Finite element models of human left ventricles (LV) of one healthy
case and one patient diagnosed with left bundle branch block (LBBB)
were fitted to clinical data consisting of LV volume, regional
longitudinal strain, and non-invasive estimates of LV
pressure. Model-data mismatch was minimised using a gradient-based
optimization technique. Regional myocardial work was assessed using
several approaches, including the area of pressure-strain loops and
the area of fiber stress-strain loops. Qualitative indices of
efficiency and quantitative measures of mechanical power density were
compared using the different approaches, including the well
established pressure-volume area.

The results show an excellent fit between measured and simulated
ventricular volumes and strains. The different methods to calculate
regional work are shown to be qualitatively similar, and they are all
able to capture regional dysfunction. However, the different
approaches are quantitatively very different, and the best agreement
between the sum of regional work contributions and the actual stroke
work is obtained using stresses and strains estimated with a finite
element model. In conclusion, mechanical work computed using a
personalized cardiac mechanics model is quantitatively similar to the
actual mechanical work estimated by the area of the pressure-volume
loop. This allows for potentially non-invasive quantitative assessment
of regional myocardial work, and would greatly improve diagnostics of
patients with regional dysfunction, such as dyssynchronous heart
failure.

@@

Typically for LBBB is that septal segments are activated early,
causing a stretch of the opposing lateral wall. The 
contraction of the late-activated lateral wall then results in a 
lengthening of the septal segments during systole, which can be
identified as wasted work. Recent studies 
\cite{vecera2016wasted} have shown that wasted septal work due to the
uncoordinated contraction is a good predictor for response to
CRT, and that the amount of wasted work reflects the potential for
recovery. In these calculations, regional work is computed as the area
of the pressure-strain loops, where regional longitudinal strain is
measured using speckle tracking echocardiography (SPE), and left
ventricular pressure (LVP) is estimated using a non-invasive method
\cite{russell2012novel}. The same approach, though with other ways of
assessing regional strain and LVP, has been used in earlier studies to
quantify regional myocardial work
\cite{tyberg1974analysis,theroux1974regional}. These calculations are
inspired by the well established correlation between pressure-volume
(PV) area and myocardial oxygen consumption \cite{suga1979total}.
%  the
% area bounded by the end-systolic PV line, the systolic limb of the PV
% loop trajectory and the volume-axis,

@@

Although the pressure-strain loop area as an estimate of regional
myocardial work has shown to be a useful metric in e.g predicting the
response to CRT, it suffers from limitations when compared to computation of 
``real'' myocardial work. Intraventricular pressure is
used as a surrogate for stress, because stress is not possible to
directly measure in an intact heart
\cite{huisman1980measurement}. Consequently, in these calculations all
segments are assumed to be subjected to the same stress, regardless of
differences in overall size, curvature or wall thickness.  
These limitations have been evaluated by Delhaas et. al
\cite{delhaas1994regional} who showed that
the area of regional fiber stress-strain loops was correlated to
regional oxygen demand in six canine left ventricles. In this study,
regional fiber stress was estimated based on assumptions of rotational
symmetry \cite{arts1991relation}, with each segment scaled according to
measured fiber strain. Fiber strain was measured at the subepicardial
layers using a two-dimensional video technique.
However, several simplifications, in particular regarding symmetry, is
made in these simplified models \cite{zhang2011comparison}, and the
measurement has to be done {\it in situ}.

@@

A framework for assessing regional estimates of myocardial
work, for which, when all regions are considered, sums up to the total
stroke work needed to eject blood to the body, is still
lacking. Furthermore, in order to be clinically useful, the framework
has to be accurate in terms of being able to replicate the mechanical
conditions observed clinically, and fast in terms of computing these
estimates within a reasonable time frame.

In this study we apply a recently developed framework
\cite{balaban2017high} for constraining a mechanical model to clinical
data. This framework employs gradient-based optimization to estimate high
dimensional parameters in order to minimize the model-data
mismatch. Moreover, the framework utilizes state-of-the-art numerical
methods which computes the gradient efficiently by solving the
corresponding adjoint equation, and thereby provides these estimates
within a reasonable time frame.

@@

Incompressibility is enforced by using a two-field variational approach
where the term
\begin{align}
\Psi_{\text{vol}} = - p (J-1), 
\end{align}
is added to the total strain energy density, and the variable $p$ is
introduced as a Lagrange multiplier.

@@

The force balance equation can be derived from the classical Cauchy
momentum equation, and with the assumption that the inertial term and
external body forces are negligible, the strong formulation yields
\begin{align}
  \nabla \cdot \FPiola &= 0, \\
  J &= 1, 
\end{align}
with $\FPiola = \frac{\partial \Psi}{\partial \F}$ being the first
Piola-Kirchhoff stress tensor.

To solve the problem we employ the Galerkin finite element method, which is based
on choosing proper finite dimensional subspaces $V_h \subset [H_0^1(\Omega)]^3$ and
$Q_h \subset L^2(\Omega)$ and solve the weak formulation of the problem. In
the Lagrangian form, the weak form reads: {\it find $(\uvec,p)
\in V_h\times Q_h$ such that for all variations $(\delta
\uvec, \delta p) \in V_h \times Q_h $ and $\left. \left( \uvec \cdot \N \right)
\right|_{\partial \Omega_{\mathrm{base}}} = 0$ }:
\begin{align}
\begin{split}
  F(\uvec, p;\delta \uvec, \delta p) =
  &\int_{\Omega} \FPiola :  \Grad \delta \uvec  \mathrm{d} V  \\
  &+\int_{\Omega} p J \F^{-T} : \Grad \ \delta \uvec   \mathrm{d} V  
  +\int_{\Omega} (J - 1) \delta p \mathrm{d} V \\
  &+\int_{\partial \Omega_{\text{base}}} k \uvec \cdot \N \mathrm{d}S
  - \int_{\partial \Omega_{\text{endo}}} \plv J \F^{-T} \N \cdot \delta \uvec \mathrm{d}S= 0.
  \label{eq:force_balance}
\end{split}
\end{align}
Here, the ventricular domain is anchored by constraining the basal
motion by a linear spring-term with stiffness $k$, and the
endocardial cavity pressure $\plv$ is enforced as a Neumann boundary
condition on the endocardium.

@@

Work can be computed by considering any of the
work conjugate stress-strain pairs \cite{holzapfel2000nonlinear}. In
the current configuration it is common to use the Cauchy stress tensor
$\Cauchy = \FPiola \F^T$ and the Almansi strain tensor $\mathbf{b} =
\frac{1}{2} \left(\I -  \F^{-T} \F^{-1} \right)$, while in
the reference configuration we use the second Piola-Kirchhoff stress
tensor $\SPiola = \F^{-1} \FPiola$, and the Green-Lagrange strain
tensor $\E = \frac{1}{2}\left(\C - \I \right)$. Thus, in the Lagrangian
description, the total mechanical work from time $t_1$ to $t_2$ in a
region $\Omega_j$ is given by
\begin{align}
   \mathcal{W}(t_1, t_2)[\Omega_j] = 
  -\int_{t_1}^{t_2} \int_{\Omega_j}\SPiola(t, \Xvec)
  : \dot{\E}(t, \Xvec) \mathrm{d}\Xvec \mathrm{d}t, 
\end{align}
where $ \dot{\E} $ is the rate of the Green-Lagrange strain tensor.
By dividing by the volume of the region of interest $| \Omega_j |$,
we obtain the average work density in that region. Note that the sign
convention used here implies that a negative strain-rate
represents positive work. To only compute the
work along a specific direction, e.g  work done by normal stress along the fibers, the full stress
and strain tensors are simply interchanged with their respective
components in that direction, and the double contraction of the tensors
is changed to a scalar multiplication.

The mechanical work needed to eject blood from time $t_1$ to
$t_2$ is given by
\begin{align}
   \mathcal{W}(t_1, t_2)[\Omega] = 
  -\int_{t_1}^{t_2} \frac{\mathrm{d}V}{\mathrm{d}t} P(t) \mathrm{d}t,
\end{align}
where $V(t)$ and $P(t)$ refers respectively to the cavity volume and pressure in
the left ventricle $\Omega$ at time $t$. If the $t_1$ and $t_2$
represent, respectively, the start and end of the cycle, then normalizing with respect to the
total ventricular wall volume  $| \Omega |$, yields a measure of
total mechanical work density (TMWD). Diving the TMWD by the total
cycle time, yields a measure of total mechanical power
density (TMPD). We note that for analysis of performance, TMPD is
a better measure than total work because it is independent of
morphology and heart rate.

\subsubsection{Quasi-static approximation}
The mechanical model described in Section
\ref{sec:mechanical_modeling} is quasi-static, meaning that the
inertial forces are assumed to be so small that the
equilibrium equation \eqref{eq:force_balance} holds at any discrete
time point. Therefore, strain rate is estimated using a
backward difference approximation \cite{wang2011myocardial}.
Suppose we have $N+1$ discrete time points $t_i, i = 0,1, \cdots N$, where $ t_0$
refers to the the reference configuration, and let $\SPiola(t, \Xvec)$
and $\E(t, \Xvec)$ be respectively the stress and strain tensors at
time $t$ and position $\Xvec$ in the reference configuration. In the
following, the full stress and strain tensors are used, but the methodology
also applies to any component of the stress and strain tensor. We define
\begin{align}
  \overline{\SPiola}(t_i,\Xvec) &= \frac{1}{2}\left( \mathbf{S}(t_i,\Xvec) + \mathbf{S}(t_{i-1},\Xvec) \right),\\
  d\E(t_i,\Xvec) &= \E(t_i,\Xvec) - \E(t_{i-1},\Xvec),
\end{align}
for $i = 1,\cdots, N$, and note that $ \mathbf{S}(t_0) = \mathbf{E}(t_0) = \mathbf{0}$.
The mechanical work in a region $\Omega_j$ from time $t_m$ to time $t_n$ can now be approximated
as 
\begin{align}
  \begin{split}
   \mathcal{W}(t_m, t_n)[\Omega_j]
  &= -\int_{t_m}^{t_n} \int_{\Omega_j}\SPiola(t, \Xvec) : \dot{\E}(t, \Xvec) \mathrm{d}\Xvec \mathrm{d}t \\
  &\approx - \sum_{i = m}^{n}\int_{\Omega_j}\overline{\SPiola}(t_i, \Xvec) : d\E(t_i, \Xvec) \mathrm{d}\Xvec
  \end{split}
\label{eq:total_work}
\end{align}
Similarly, the mechanical work computed from the pressure-volume loop area can be estimated as  
\begin{align}
  \begin{split}
  \mathcal{W}(t_m,t_n)[\Omega]
  &= - \int_{t_m}^{t_n} \frac{\mathrm{d}V}{\mathrm{d}t} P(t) \mathrm{d}t \\
  &\approx -\sum_{i= m}^{n}  (V(t_i) - V(t_{i-1})) \cdot \frac{P(t_i) + P(t_{i-1})}{2}.
  \end{split}
\end{align}

@@

During systole, most segments shorten and perform positive work
($\mathcal{W}_{\mathrm{pos}}$). However some segments might elongate,
resulting in energy loss, and we say that these segments do negative work
($\mathcal{W}_{\mathrm{neg}}$). Positive work is identified as the points with
$d\E_{\ef}(t) < 0$ (shortening), while negative work is identified as points with
$d\E_{\ef}(t) > 0$ (lengthening).
A local measure of the energetic efficacy of each segment is the ratio between the
negative and positive work, an index known as the segmental wasted
work ratio (WWR) \cite{russell2013assessment},
\begin{align}
  \mathrm{WWR} = \frac{|\mathcal{W}_{\mathrm{neg}}|}{\mathcal{W}_{\mathrm{pos}}}
\end{align}
Another measure of efficiency, formulated by Niederer et al
\cite{niederer2009role} is the ratio of positive work and the total
work
\begin{align}
  \eta = \frac{\mathcal{W}_{\mathrm{pos}}}{\mathcal{W}_{\mathrm{pos}} + |\mathcal{W}_{\mathrm{neg}}|}.
  \label{eq:efficiency}
\end{align}

The two indices are related by $\eta = \left( 1 +
  \mathrm{WWR} \right)^{-1}$, and we will therefore only report values
of the efficiency \eqref{eq:efficiency}, for which a value close to 1.0 indicate that most
of the performed work is positive. Since we expect work during systole
to be the main contribution to the overall ejection of blood, the
index is computed by only including work done from mitral valve closing to
mitral valve opening \cite{russell2013assessment}.

% Note that with the current sign-convention, negative work is by
% definition negative, which is why we take the absolute value of the
% negative work.

@@

In this study 4D echocardiographic images of one patient diagnosed
with left bundle branch block (LBBB) and one healthy volunteer was
acquired using a GE Vingmed E9 machine (GE Healthcare Vingmed, Horten,
Norway). 
The analysis of the echocardiographic images was conducted with the
GE software package EchoPac version 202. Longitudinal strain was
measured via speckle tracking echocardiography (SPE) in the 17 AHA segments
\cite{cerqueira2002standardized}. However, the apical segments for both
subjects, as well as the inferior and posterior basal segment for the
normal case, were excluded due to poor quality.

@@

The initial guesses for the optimization in this phase were set to
zero in the first iteration, and in subsequent iterations, the initial
guesses were set to the optimized values found in the previous iteration.

\section{Results}
For the normal and LBBB case described above, we ran the
full data assimilation pipeline on two meshes, where one was a refined version of the
other. For the normal case, the original and refined mesh had
respectively 2569 and 5258 elements, while the LBBB case
had 2572 elements in the original mesh and 5114 elements in the
refined mesh. Results on the original and refined geometries gave
close to identical results, and we will therefore only present the
results using the coarsest geometry.

@@

During the estimation of the active and passive parameters outlined in Section
\ref{sec:active_param_estim} and \ref{sec:passive_param_estim}
respectively, the optimization was terminated if the difference between two
successive iterations was less then $10^{-6}$ or the number of
iterations exceeded 100. For the material parameter, an upper and lower
bound was set to $50$ and $0.05$ kPa respectively, while for the active
parameter these bounds were set to $0.4$ and $0.0$, allowing for up to
40 $\%$ active shortening of the myofibers.

@@

The total mechanical power density (TMPD) is also computed using four
different approaches: 1) area of the pressure-volume loop normalized
to ventricular wall volume, 2) using the full stress and strain tensors,
3) using the fiber component of the stress and strain tensors, and 4)
using global longitudinal strain and estimated pressure. Table
\ref{tab:total_mechanical_power} show these values for the two
subjects, in units kW m$^{-3}$. In the calculations of pressure volume
area, the simulated volume is used, with the unloaded volume and zero
pressure included as reference point.

@@

A qualitative comparison of the pressure-strain loops and fiber
stress-strain loops are illustrated in Figure
\ref{fig:regional_work_indices}, where we plot the regional efficiency
\eqref{eq:efficiency} using the two measures of work. We note that the
efficiencies computed using the fiber stress-strain loops are
lower than for the pressure-strain loops, but show the same trend, i.e
that regions with reduced efficiency coincide.

@@

The values in Table \ref{tab:total_mechanical_power} indicate that the
total mechanical work performed in the direction of the fibers is
approximately 60 $\%$ of the 
total mechanical work computed by taking into account the full stress
and strain tensor (LBBB: 60.5 $\%$, normal:58.7 $\%$). Of course, the
amount of work performed in transverse direction of the fibers will depend upon the
chosen form of the active deformation gradient in
\eqref{eq:active_strain_Fa_gjerald}. In the active stress approach on 
the other hand, where the total Cauchy stress tensor is additively decomposed into
active and passive contributions, it is common to add an active stress
in the transverse direction, which is also reported
experimentally \cite{lin1998multiaxial}. By assuming
a volume preserving active deformation gradient, the amount of
work in the cross fiber direction as computed here is about $40 \%$, a
value that has been used in other modeling studies using the active
stress approach \cite{sun2009computationally}.

The values of regional TMPD in the mid segments presented
in Figure \ref{fig:regional_power_density} show that the septal
segments for the LBBB patient actually consume energy and yield a
negative TMPD. On the other hand, the healthy heart shows positive TMPD
in all segments and generally higher values than the LBBB heart.
Since the TMPD computed using stress and strain derived from the
underlying constitutive equation agrees very well with the TMPD
computed from the PV area there is reason to believe that the regional
TMPD reflects the actual regional performance, and future studies should
investigate how well these values correlate with regional myocardial
oxygen consumption.

The TMPD computed using the pressure and global longitudinal strain
only accounts for a small fraction of the total mechanical power.
This indicates that quantitatively this is not a good measure of
myocardial work. However, the area of the pressure-strain loops have shown to be a good metric
of regional efficiency and a promising index for predicting response
to CRT \cite{vecera2016wasted}. From Figure
\ref{fig:regional_work_indices} we see the same trends in terms of efficiency and wasted
work can be seen regardless of how work is computed, namely that early
activated segments are less efficient and have a higher wasted work
ratio than the other segments for the LBBB case. Therefore, relative
comparison of segments using pressure-strain loops seems to give the
same picture as when using the fiber stress-strain loops. 
The pressure strain loops for the LBBB case all have an
anti-clockwise orientation, indicating generation of work in all
segments. On the other hand, the fiber stress-strain loops in the
septal segments show a clockwise orientation which is evidence of
work usage. In other words, the two approaches show the same
trend, but the effect of the wasted work is magnified in the fiber
stress-strain loops.

@@

The optimization problem during the active phase is highly
under-constrained, which is why we apply a Tikhonov regularization to
restrict the parameter space, so that smooth value of the controls are
favored in the optimization. While the motivation for this choice is
numerical stability, it may be debated whether this is a valid
assumption from a physiological point of view, especially in the case
of ventricular dyssynchrony.

Experiments have shown both a visco-elastic and an orthotropic
behavior of the myocardium
\cite{dokos2002shear,sommer2015biomechanical}.
Visco-elasticity would certainly play a role in myocardial work estimation,
as the energy is preserved in a hyperelastic material while some energy dissipate when
visco-elasticity is taken into account. However, the effect of
visco-elasticity and orthotropy is not modeled here, and remains a
subject for future studies.
